%% ----------------------------------------------------------------
%% Thesis.tex -- MAIN FILE (the one that you compile with LaTeX)
%% ---------------------------------------------------------------- 

% Set up the document
\documentclass[a4paper, 12pt, oneside]{Thesis}  % Use the "Thesis" style, based on the ECS Thesis style by Steve Gunn
\graphicspath{Figures/}  % Location of the graphics files (set up for graphics to be in PDF format)

% Include any extra LaTeX packages required
\usepackage[square, numbers, comma, sort&compress]{natbib}  % Use the "Natbib" style for the references in the Bibliography
\usepackage{verbatim}  % Needed for the "comment" environment to make LaTeX comments
\usepackage{vector}  % Allows "\bvec{}" and "\buvec{}" for "blackboard" style bold vectors in maths
\usepackage[utf8]{inputenc}
\usepackage{aas_macros}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{soul}
\hypersetup{colorlinks=true,allcolors=teal}  % Colours hyperlinks in blue, but this can be distracting if there are many links.

% \hypersetup{colorlinks=true,allcolors=teal}

\newcommand{\doi}[1]{\href{https://doi.org/#1}{doi:#1}}
\newcommand{\Hi}{\textsc{Hi}}
\newcommand{\mHi}{\ensuremath{m_{\Hi}}}

\newcommand{\p}{\ensuremath{\partial}}


\newcommand{\Msun}{\ensuremath{M_{\odot}}}
\newcommand{\Mh}{\ensuremath{h^{-1}M_{\odot}}}
\newcommand{\Mhsq}{\ensuremath{h^{-2}M_{\odot}}}
\newcommand{\Mpch}{\ensuremath{h^{-1}{\rm Mpc}}}
\newcommand{\kpch}{\ensuremath{h^{-1}{\rm kpc}}}
\newcommand{\kms}{\ensuremath{{\rm km\,s}^{-1}}}
\newcommand{\msq}{\ensuremath{{\rm \,m\,s}^{-2}}}

\newcommand{\avg}[1]{\ensuremath{\left\langle \,#1\, \right\rangle}}
\newcommand{\e}[1]{\ensuremath{{\rm e}^{#1}}}

\newcommand{\der}{\ensuremath{{\rm d}}}
\newcommand{\Der}{\ensuremath{{\rm D}}}
\newcommand{\dir}{\ensuremath{\delta_{\rm D}}}

\newcommand{\erfc}[1]{\ensuremath{{\rm erfc}\left(#1\right)}}
\newcommand{\erf}[1]{\ensuremath{{\rm erf}\left(#1\right)}}

\newcommand{\eqn}[1]{equation~\eqref{#1}}
\newcommand{\eqns}[1]{equations~\eqref{#1}}
\newcommand{\eqnref}[1]{equation~\eqref{#1}}
\newcommand{\eqnsref}[1]{equations~\eqref{#1}}
\newcommand{\ph}[1]{\phantom{#1}}
\newcommand{\figref}[1]{Figure~\ref{#1}}
\newcommand{\tabref}[1]{Table~\ref{#1}}
\newcommand{\secref}[1]{Section~\ref{#1}}
\newcommand{\appnref}[1]{Appendix ~\ref{#1}}

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\Cal}[1]{\ensuremath{\mathcal{#1}}}

\newcommand{\AP}[1]{\emph{\color{blue}[AP: #1]}}
\newcommand{\PV}[1]{\emph{\color{violet}[PV: #1]}}
\newcommand{\red}[1]{\textcolor{red}{#1}}

%% ----------------------------------------------------------------
\begin{document}
\frontmatter      % Begin Roman style (i, ii, iii, iv...) page numbering

% Set up the Title Page
\title  {Interplay of galaxy formation and the evolution of dark matter haloes in the cosmic web}
\authors    {\texorpdfstring{\href{premv@iucaa.in}{Premvijay Velmani}}{Premvijay Velmani}}
\addresses  {\groupname\\\deptname\\\univname}  % Do not change this here, instead these must be set in the "Thesis.cls" file, please look through it instead
\date       {\today}
\subject    {}
\keywords   {}

\maketitle
%% ----------------------------------------------------------------

\setstretch{1.5}
\mainmatter	

\chapter*{Synopsis}

\section*{Background}

Dark matter is inferred exclusively through its gravitational effects to make up over 80\% of all the matter in the Lambda-cold dark matter ($\Lambda$CDM) model of the Universe. Tiny fluctuations in the initial density field cause them to clump into gravitationally collapsed structures called haloes \citep[][]{1974ApJ...187..425P,2002PhR...372....1C}. These haloes are fundamental to understanding cosmology as they form the building blocks of the large-scale structure and also reveal insights into the particle physics of dark matter. Dark matter haloes have been extensively studied in gravity-only scenarios both in isolated systems and over cosmological volumes. For example, the haloes identified in cosmological $N$-body simulations on $\Lambda$CDM cosmology are known to be triaxial \citep[][]{1988ApJ...327..507F}, and their sphericalised mass profiles are found to have a universal form \citep{1996ApJ...462..563N,1997ApJ...490..493N,2010MNRAS.402...21N}.

Dark matter haloes also provide the primary environment for the formation and evolution of galaxies \citep[e.g.,][]{wr78}, making them crucial for understanding astrophysical processes. Gas, through various non-gravitational baryonic interactions, can cool and condense towards the halo center \citep{1988MNRAS.234..459S,1998MNRAS.295..319M}, eventually producing stars and other astrophysical bodies. However, the complexity and incomplete knowledge of these baryonic processes, along with their large dynamic range, make it impossible to study galaxy formation directly from first principles using cosmological simulations.

One approach to studying galaxy formation is to use semi-analytic techniques, modeling its formation within the halo based on properties known from gravity-only simulations. These semi-analytic models have various free parameters that need to be constrained typically by observations \citep{2015ARA&A..53...51S}. The formation of galaxies driven by the baryonic interactions within these haloes also affect the spatial distribution of dark matter through gravitational coupling. This response of a halo's dark matter content to the baryons it hosts must be accounted for to understand the coupled evolution of haloes and galaxies.

The state-of-the-art approach to galaxy formation involves numerical hydrodynamical simulations in cosmological volumes, using subgrid recipes to model baryonic physics at unresolved scales. Although some baryonic processes are included only as effective models with parameters constrained by observations, this approach is more realistic than semi-analytical models for making predictions. This method also incorporates the backreaction of galaxies on dark matter haloes, though it is computationally challenging. Modern cosmological simulations like EAGLE \citep{2015Schaye_EAGLE} and IllustrisTNG \citep{2018MNRAS.480.5113M} were performed on high-performance clusters over several months.

Understanding the mutual effects of dark matter halo formation and galaxy formation on each other is relevant for various problems. For example, halo shapes respond to galaxy formation \citep{2010MNRAS.407..435A,2021MNRAS.501.5679C}, which is important for weak-lensing studies \citep{2021A&A...647A.185G}. Similarly, galaxy formation changes the dark matter density profile of the halo, affecting the galaxy's rotation curve. However, previous works have not reached a consensus on the backreaction of galaxy formation on halo properties in simulations, making it a pressing open problem.

From a cosmological viewpoint, these `baryonic' effects are often a nuisance in parameter inference; e.g., the effects of feedback from active galactic nuclei (AGN) can be degenerate with the effects of a massive neutrino species or a thermally produced ‘warm’ dark matter candidate at Fourier scales $k\sim1\,h\,{\rm Mpc}^{-1}$ (see, e.g., \cite{2019Chisari_etal_Baryfeedback,2020AricoAnguloetal_baryonifi}). The response of dark matter (DM) to baryons must be quantified and distinguished from observational effects of (non-standard) primordial physics. While the former problem has a long history \cite{1986Blumenthal,2004Gnesin_etal,2005SellwoodMcGaugh,2006Gustafsson_FS,2010Abadi_NFBS,2010DuffySchaye_etal,2010PedrosaTissera_etal,2010TisseraWhite_etal,2019ArtalePedrosa_etal,2022ForouharMoreno_etal,2023Velmani&Paranjape}, the latter has only recently begun to be studied in detail \cite{2011TeyssierMMDM,2015SchneiderTeyssier,2015Mead_PHJH,2020AricoAnguloetal_baryonifi,2021AricoAnguloetal_baryonifi,2023EuclidCastro_etal}.

\section*{Thesis Statement}

In this thesis, we study the interplay between galaxy formation and its host halo evolution, focusing on changes in the radial distribution of dark matter in response to galaxy formation and evolution. We build a comprehensive understanding using both hydrodynamical simulations of realistic galaxies in cosmological volumes and more tractable self-similarly evolving systems of individual galaxies with their host haloes. This unified picture of the role of various galactic astrophysical processes in mediating the response of dark matter haloes also provides useful timescales for predicting dynamical changes in radial mass profiles due to astrophysical processes.

\section*{Overview}

\subsection*{Chapter 1}
This chapter introduces the background material and the primary objectives of the thesis, along with a review of existing knowledge and the relevance of this work in the broader context of ongoing research in physics.

\subsection*{Chapter 2}
We describe the simulation data and techniques primarily employed in this thesis. Currently, the most robust technique to understand the consequences of gas assembly and galaxy formation on dark matter structure is the use of high-resolution cosmological hydrodynamical (zoom) simulations, using `sub-grid' recipes for modeling small-scale astrophysics such as feedback from stellar/supernovae activity or the effects of active galactic nuclei (AGN). In this thesis, we mainly use publicly available data from such cosmological simulations of galaxies performed as part of large collaborative projects, namely IllustrisTNG \cite{2019ComAC...6....2N}, EAGLE \cite{2015MNRAS.446..521S}, and CAMELS \cite{CAMELS_presentation,CAMELS_DR1}. This chapter provides a brief description of these simulations and the identification of halo objects in their respective cosmological volumes, focusing on details relevant to this thesis.

We also develop a matching algorithm to assign each halo in the full hydrodynamic simulations a matched 'partner' halo in a collisionless, gravity-only simulation performed using the same initial random fluctuations. We demonstrate its robustness, as comparing these hydrodynamically simulated haloes against their partner haloes is key to investigating the simulated backreaction of galaxies on dark matter haloes.

Early work modeled halo response using adiabatic invariants within individual haloes \citep[][]{osti6457593,1984MNRAS.211..753B,1986ApJ...301...27B,1987ApJ...318...15R}. Using simplifying assumptions such as spherical symmetry, no shell crossing, angular momentum conservation with circular orbits for dark matter particles, a simple formula was derived to quantify the \emph{adiabatic relaxation} of the radial distribution of dark matter in terms of the radial distribution of baryons. This idealized model of adiabatic relaxation \citep[][]{1986ApJ...301...27B} was found to be inaccurate in various simulations \citep[see, e.g.,][]{2004ApJ...616...16G,2006PhRvD..74l3522G,2010MNRAS.402..776P,2010MNRAS.406..922T,2010MNRAS.405.2161D,2010MNRAS.407..435A,2011MNRAS.414..195T,2016MNRAS.461.2658D,2019A&A...622A.197A,2022MNRAS.511.3910F}. This led to developing models that are direct extensions of the idealized model with additional parameters constrained by the simulated response \citep{2010MNRAS.407..435A,2004ApJ...616...16G,2006PhRvD..74l3522G,2010MNRAS.405.2161D}.

We characterize halo response within this quasi-adiabatic relaxation framework, primarily through the relation between relaxation ratios and enclosed mass ratios, quantifying changes in the sphericalised mass profiles of dark matter and baryons, respectively. The chapter concludes with a description and demonstration of the techniques used in computing these quantities in simulated pairs of haloes.



\subsection*{Chapter 3}

In this chapter we perform a systematic, statistical study of the dark matter response to galaxies in high-resolution hydrodynamical simulations of realistic galaxies in cosmological volumes from IllustrisTNG and EAGLE projects at the present epoch.
% 
% and quantify it using simple analytical forms, including the sensitivity of this response to halo-centric distance and halo and galaxy properties. To this end, we use the publicly available suites of simulations from the IllustrisTNG and EAGLE projects. 
% 
% 
% In this chapter, we delve into the response of dark matter within haloes to the baryonic processes, primarily focusing on the results obtained from the simulations. 
% 
We explore this response across a wide range of halo masses, spanning four orders of magnitude, and over various halo-centric distances at redshift $(z simeq 0)$.

We establish that the dark matter's response to baryons cannot be accurately described by the commonly used quasi-adiabatic relaxation models, which typically relate the relative change in radius (\(r_f/r_i-1\)) of a dark matter shell solely to the relative change in their enclosed masses (\(M_i/M_f-1\)). Instead, our findings reveal that the relaxation response (\(r_f/r_i\)) also explicitly depends on the halo-centric distance (\(r_f/R_{\rm vir}\)), a dependency that is consistent across different halo masses and between the IllustrisTNG and EAGLE simulations.

Our study also identifies an unmodeled effect likely driven by feedback-related outflows. This effect shows that even shells with no overall change in the enclosed mass (\(M_i/M_f \simeq 1\))  can exhibit relaxation of its radius (\(r_f/r_i < 1\)). This suggests that existing models need to incorporate this additional parameter to more accurately capture the complex interplay between dark matter and baryonic processes.

We present a new, physically motivated model that incorporates these findings, offering a better fit for the dark matter response over the examined range of halo masses and distances. This model introduces an additional parameter to account for feedback-induced offsets, providing a more comprehensive description of the dark matter relaxation process.

Moreover, we investigate the dependence of this response on various halo and galaxy properties beyond total mass. We find that halo concentration and star formation rate play significant roles in influencing the dark matter response. These insights are crucial for improving semi-analytical models and for applications such as baryonification schemes in emulating the small-scale power spectrum and modeling the rotation curves of galaxies.





% Physically, one expects that the overall response of the halo is mediated by a combination of different astrophysical processes that occur in the galaxy. Feedback processes are known to reduce the contraction of the halo significantly; e.g.,
% supernova-driven winds can completely transform the inner density profile of the dark matter halo
% \citep[][]{1996MNRAS.283L..72N}. This may be the key in reconciling the observation of dark matter cores at the center of various galaxies with the cuspy haloes found in gravity-only $\Lambda$CDM simulations \citep[see][for a review]{2014Natur.506..171P}.
% However, such feedback effects do not always produce dark matter cores from cusps, rather, this can depend on the amount of gas ejected, the mass loss time scale and the frequency of starburst events 
% (see, e.g., \citealp{2011ApJ...736L...2O,2014ApJ...793...46O,2012MNRAS.421.3464P}, and also \citealp{bfln18}).
% In massive haloes hosting galaxy groups or clusters, while the formation of powerful AGN in the central galaxy can strongly suppress star formation,
% it can still significantly reduce the adiabatic contraction of the halo \citep[][]{2011MNRAS.414..195T}.
% Moreover, the fluctuation in gravitational potential due to such feedback can expel the dark matter from the inner halo producing inner cores \citep[][]{2012MNRAS.422.3081M}.


% In our earlier work \cite{2023Velmani&Paranjape} we quantified this response in the IllustrisTNG \cite{2018TNG_Pillepich_etal}
% and EAGLE \cite{2015Schaye_EAGLE} %
% simulation suites by studying the mass profiles of matched pairs of halos in the hydrodynamical and gravity-only runs of the simulation, and described the response in the language of quasi-adiabatic relaxation \cite{2011TeyssierMMDM}. %
% In particular, we showed that the original quasi-adiabatic prescription accurately describes the relaxation of halos in simulations provided we include the halo-centric distance in units of the halo radius as an additional parameter. We further provided convenient fitting forms for this parametrised relaxation at $z=0$ and explored its dependence on various halo and galaxy properties such as halo concentration, stellar mass, star formation rate, etc.



\subsection*{Chapter 4}

In this chapter, we investigate the influence of astrophysical modeling on the relaxation response of dark matter haloes at different epochs, specifically focusing on \( z=0 \) and \( z=1 \). The analysis is divided into three main parts, each shedding light on the role of various astrophysical processes in shaping the dark matter content of haloes.

\subsubsection*{1. Early Epoch in IllustrisTNG Simulations}
We begin by examining the relaxation response at an earlier redshift (\( z=1 \)) using the IllustrisTNG simulations. Our study reveals that dark matter relaxation tends to be stronger at \( z=1 \) compared to \( z=0 \). We assess this using four distinct halo samples, which highlight the variations in relaxation across different halo masses. Notably, we observe that cluster-scale haloes at \( z=1 \) show significant relaxation that becomes a function of the change in the enclosed mass, a contrast to similar haloes at the present epoch.

Our findings suggest that the relaxation relation at this earlier epoch can also be modeled effectively using the locally linear model developed in the chapter 3, demonstrating the robustness of this approach in capturing the dark matter response across redshifts. Moreover, the parameters of the radially dependent relaxation are found to be more universal across a much wider range of masses at \( z=1 \). For example the progenitors of even the most massive clusters are well characterised by the simple three parameter model of relaxtion that was developed with a focus on galactic scale haloes at \( z=0 \)

\subsubsection*{2. Variation in Astrophysical Feedback Using CAMELS Simulations}
Next, we explore the variations in astrophysical feedback strengths in the IllustrisTNG model of the simulations of the galaxies. This analysis is performed on a large number of small volume simulations provided by the CAMELS project that varies four different feedback parameters. These four parameters  include two for the stellar feedback and other two for the AGN feedback. Our analysis shows that the parameters controlling the energy flux of the feedbacks had significant impact on the relaxation of dark matter at different epochs. However, the parameters controling the speed and burstiness of the feedback have only negligible effect on the halo relaxation response.

We also see that the variation in the stellar feedback strengths has larger  impact among dwarf galaxy scale haloes while variation in the AGN feedback parameters have stronger influence on the milkyway scale haloes. In all those haloes the relaxation offset in the outer well resolved regions is stronger at present epoch than earlier epoch (\(z=1\)) unlike results from the more the inner regions explored in the IllustrisTNG simulation presented in the first part of this chapter. 

The stronger implementation of AGN feedback tends to make stronger relaxation at both \( z=0 \) and \( z=1 \) in the outer regions of the haloes. However in the slightly inner regions, it makes the relaxation offset weaker at \( z=0 \) and stronger at \( z=1 \). We interpret this as a consequence of the overall reduction in the total feedback at \( z=0 \) due to the suppression of star by higher AGN feedbacks in the past. These results highlight the significance of feedback mechanisms in building a physical understanding of the dark matter halo relaxation.

\subsubsection*{3. Role of Astrophysical Models in the EAGLE Simulations}
Finally, we assess the impact of different astrophysical models in the EAGLE simulations. Supernova feedback strengths show a similar trend that was seen with the CAMELS simulation. In addition, we find that the gas equation of state to have the strongest effect on the relaxation response of the dark matter especially among the haloes hosting dwarf galaxies. 

Overall, this chapter underscores the intricate relationship between baryonic processes and dark matter halo relaxation, illustrating the variations that arise due to different astrophysical models and redshifts.



\subsection*{Chapter 5}



\subsection*{Chapter 6}

The most realistic description of halo response or `relaxation' in the presence of baryons is provided by hydrodynamical simulations that include calibrated effects of star formation, radiative cooling, feedback due to stellar winds and supernovae as well as the triggering of AGN activity in super-massive black holes and its feedback on star formation \cite{2015Schaye_EAGLE,2018TNG_Pillepich_etal}. 


Physically, however, the mechanisms driving this halo response are not entirely clear. In principle, this question can also be addressed using hydrodynamical simulations by systematically varying the physical prescriptions defining different processes, particularly the observationally uncertain ones associated with stellar and especially AGN feedback. This, however, is a very computationally expensive exercise requiring multiple runs of expensive simulations \cite{2015Schaye_EAGLE,2021camels_presentation}.
We will report the results of such an exercise using the CAMELS simulation suite \cite{2022camels_data_release1} in a separate publication. In the present work, we investigate whether we can gain some physical insights into the problem of halo relaxation due to baryonic backreaction using a simplified, spherically symmetric toy model of galaxy evolution.

Our key assumption will be that the model is `self-similar', so that spatial variations are related to temporal variations through a scale radius that evolves over time. The self-similar assumption offers a powerful tool to simplify the problem and render it more tractable, by converting nonlinear coupled partial differential equations into ordinary differential equations. The self-similar model was pioneered by Fillmore and Goldreich (1984) \cite{1984FillmoreGoldreich} and Bertschinger (1985) \cite{1985Bertschinger}. While Bertschinger devised spherical self-similar solutions for the collapse of dark matter and shocked gas within the Einstein-de Sitter (EdS) universe, under the assumption of a constant accretion rate, Fillmore and Goldreich explored self-similar solutions for the collapse of dark matter from various initial mass distributions, expanding their purview to include not only spherical but also cylindrical and planar self-similar collapses. Bertschinger (1989) \cite{1989Bertschinger} delved into self-similar cooling flows of gas, focusing on the inner regions of galaxies. Owen et al. (1998) \cite{1998OwenWeinberg_etal} derived a set of cooling functions that guarantee self-similar evolution by ensuring that the cooling time-scale of an object with a characteristic clustering mass remains a constant fraction of the Hubble time. Abadi et al. (2000) \cite{2000Abadi_etal_SelfSimCool} studied the self-similar accretion of gas with radiative cooling in the EdS Universe. Shi (2016a) \cite{2016ShiDMLamCDM} generalised self-similar models to Lambda-cold dark matter ($\Lambda$CDM) models, focusing on the outer profile and, notably, the outermost caustic or `splashback radius' of dark matter collapse which has gained recent popularity \cite{2014DiemerKrastov,2014AdhikariDalalChamberlain,2018Changetal_DES_splashback}.
In a complementary study, Shi (2016b) \cite{2016ShiICM} probed the self-similar accretion of shocked gas, dissecting its behavior with respect to accretion rates and revealing correlations between the shock radius of gas and the dark matter's splashback radius.


The present work draws upon these foundations but incorporates two novel additions: (i) a self-similar cooling of the gas and (ii) the formation of a pseudo-disk of gas along with the self-similar collapse of the dark matter halo. Unlike the work of \cite{1989Bertschinger}, where the cooling scale was set by a cooling radius and was primarily applicable to the innermost regions, in our work we set the cooling scale relative to the turn-around radius. This allows us to keep intact the global self-similarity of the model, while allowing for a range of possible amplitudes for the cooling rate. Our inspiration for including the formation of a pseudo-disk in our spherical approach comes from spherical hydrodynamical simulations that have previously explored the influence of cooling on shock-heated gas, tracing its subsequent accretion onto central structures resembling galaxy disks \cite{2006Dekel&Birnboim}. These simulations have also revealed a `cold mode' of gas accretion that occurs in low mass halos, wherein the gas directly accretes onto an inner disk without shocking. While these results are consistent with those of full hydrodynamical simulations \cite{2005Keres_KWD},
the main relevance to us is the fact that the inner disk radius undergoes an approximately self-similar evolution, following the behavior of the turnaround radius, thus rendering it amenable to our self-similar modeling.

\emph{Importantly, we self-consistently solve for the evolution of the gas and the dark matter simultaneously using a novel iterative technique.} This approach allows us to reliably estimate the relaxation of the dark matter due to baryonic backreaction in this self-similar model. The inclusion of cooling and the formation of a pseudo-disk also brings our self-similar model of galaxy formation substantially closer to the more realistic results of hydrodynamical simulations. We leave the incorporation of the formation of stars and a central black hole, and the associated stellar and AGN feedback, to future work.


\subsection*{Chapter 7}
We conclude this thesis with a brief summary of the key findings and their applications. We also discuss the future directions to this work in this chapter.


\section*{Organization of the thesis}

In the first chapter of this thesis, we review the background material and then define the primary objective of this thesis, along with its relevance to the broader context.

In the second chapter, we demonstrate the specific techniques we employed in studying simulated response of the dark matter halo to the galaxy formation and evolution in simulations of cosmological volumes.

In the third chapter, we present our findings from state-of-the-art galaxy formation simulations, IllustrisTNG and EAGLE simulations, characterizing the dark matter response in a wide variety of haloes. We present novel fitting functions that accurately describe this relaxation response, revealing an additional dependence on halo-centric distance and highlighting the significant role of star formation-related feedback processes. 

In the fourth chapter, we explore the role of astrophysical modeling in producing the halo response at different epochs in simulations. We find the gas equation of state to have significant influence in EAGLE haloes, and we also quantify the strong influence of AGN and stellar feedback in CAMELS simulations. Our results also show a more universal relaxation response at an earlier epoch $(z=1)$ compared to $z=0$. These findings are applicable to semi-analytical tools for modeling galactic and large-scale structures. 

In the fifth chapter, we uncover the causal connections between star formation activities, feedback processes, and halo relaxation. Through time-series analyses we provide new insights into the immediate and delayed effects across different halo regions. Our estimates of the time-scales can potentially improve the description of halo profiles in existing baryonification schemes and semi-analytical galaxy formation models. 

In the sixth chapter, we develop a spherical self-similar model for galaxy formation that simultaneously and self-consistently solves for the evolution of gas and dark matter, producing pseudo galaxy disks within the halo. This complementary approach offers a framework to rapidly explore the sensitivity of halo response to various astrophysical processes. We find quasi-adiabatic halo relaxation similar to full non-linear simulations, with similarly strong trends over the gas equation of state.

In the seventh chapter, we summarize the key findings in this thesis and discuss the future outlook.

% \section*{Publications}
\section*{Publications included}

\noindent
1. Velmani P., Paranjape A., 2023, "The quasi-adiabatic relaxation of haloes in the IllustrisTNG and EAGLE cosmological simulations", \textit{Monthly Notices of the Royal Astronomical Society}, 
\textbf{520}(2):2867-2886. 
\doi{10.1093/mnras/stad297}
\href{https://ui.adsabs.harvard.edu/abs/2023MNRAS.520.2867V}{https://ui.adsabs.harvard.edu/abs/2023MNRAS.520.2867V}

\vspace{0.5cm}

\noindent
2. Velmani P., Paranjape A., 2024, "Dynamics of the response of dark matter halo to galaxy evolution in IllustrisTNG", \textit{arXiv e-prints}, 
\href{https://ui.adsabs.harvard.edu/abs/2024arXiv240708030V}{arXiv:2407.08030}. 
\doi{10.48550/arXiv.2407.08030}

\vspace{0.5cm}

\noindent
3. Velmani P., Paranjape A., 2024, "A self-similar model of galaxy formation and dark halo relaxation", \textit{Journal of Cosmology and Astroparticle Physics}, 
\textbf{2024}(5):080. 
\doi{10.1088/1475-7516/2024/05/080}
\href{https://ui.adsabs.harvard.edu/abs/2024JCAP...05..080V}{https://ui.adsabs.harvard.edu/abs/2024JCAP...05..080V}

\vspace{0.5cm}

\noindent
4. Velmani P., Paranjape A., 2024, "Role of astrophysical modeling on dark matter halo relaxation response at redshifts z = 0 and z = 1", \textit{in prep.}


% % Include the bibliography
% \bibliographystyle{plain} % You can choose another style such as plain, abbrv, unsrt, etc.
% \bibliography{publications} % 'publications' is the name of your .bib file without the extension



\bibliographystyle{abntex2-num}
\bibliography{Bibliography,references-AdiabRelxn1,references-SelfSimRelxn}


\end{document}  % The End
%% ----------------------------------------------------------------