%% ----------------------------------------------------------------
%% Thesis.tex -- MAIN FILE (the one that you compile with LaTeX)
%% ---------------------------------------------------------------- 

% Set up the document
\documentclass[a4paper, 12pt, oneside]{Thesis}  % Use the "Thesis" style, based on the ECS Thesis style by Steve Gunn
\graphicspath{Figures/}  % Location of the graphics files (set up for graphics to be in PDF format)

% Include any extra LaTeX packages required
\usepackage[square, numbers, comma, sort&compress]{natbib}  % Use the "Natbib" style for the references in the Bibliography
\usepackage{verbatim}  % Needed for the "comment" environment to make LaTeX comments
\usepackage{vector}  % Allows "\bvec{}" and "\buvec{}" for "blackboard" style bold vectors in maths
\usepackage[utf8]{inputenc}
\usepackage{aas_macros}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\usepackage{soul}
\hypersetup{colorlinks=true,allcolors=teal}  % Colours hyperlinks in blue, but this can be distracting if there are many links.

% \hypersetup{colorlinks=true,allcolors=teal}

\newcommand{\Hi}{\textsc{Hi}}
\newcommand{\mHi}{\ensuremath{m_{\Hi}}}

\newcommand{\p}{\ensuremath{\partial}}


\newcommand{\Msun}{\ensuremath{M_{\odot}}}
\newcommand{\Mh}{\ensuremath{h^{-1}M_{\odot}}}
\newcommand{\Mhsq}{\ensuremath{h^{-2}M_{\odot}}}
\newcommand{\Mpch}{\ensuremath{h^{-1}{\rm Mpc}}}
\newcommand{\kpch}{\ensuremath{h^{-1}{\rm kpc}}}
\newcommand{\kms}{\ensuremath{{\rm km\,s}^{-1}}}
\newcommand{\msq}{\ensuremath{{\rm \,m\,s}^{-2}}}

\newcommand{\avg}[1]{\ensuremath{\left\langle \,#1\, \right\rangle}}
\newcommand{\e}[1]{\ensuremath{{\rm e}^{#1}}}

\newcommand{\der}{\ensuremath{{\rm d}}}
\newcommand{\Der}{\ensuremath{{\rm D}}}
\newcommand{\dir}{\ensuremath{\delta_{\rm D}}}

\newcommand{\erfc}[1]{\ensuremath{{\rm erfc}\left(#1\right)}}
\newcommand{\erf}[1]{\ensuremath{{\rm erf}\left(#1\right)}}

\newcommand{\eqn}[1]{equation~\eqref{#1}}
\newcommand{\eqns}[1]{equations~\eqref{#1}}
\newcommand{\eqnref}[1]{equation~\eqref{#1}}
\newcommand{\eqnsref}[1]{equations~\eqref{#1}}
\newcommand{\ph}[1]{\phantom{#1}}
\newcommand{\figref}[1]{Figure~\ref{#1}}
\newcommand{\tabref}[1]{Table~\ref{#1}}
\newcommand{\secref}[1]{Section~\ref{#1}}
\newcommand{\appnref}[1]{Appendix ~\ref{#1}}

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\Cal}[1]{\ensuremath{\mathcal{#1}}}

\newcommand{\AP}[1]{\emph{\color{blue}[AP: #1]}}
\newcommand{\PV}[1]{\emph{\color{violet}[PV: #1]}}
\newcommand{\red}[1]{\textcolor{red}{#1}}

%% ----------------------------------------------------------------
\begin{document}
\frontmatter      % Begin Roman style (i, ii, iii, iv...) page numbering

% Set up the Title Page
\title  {Interplay of galaxy formation and the evolution of dark matter haloes in the cosmic web}
\authors    {\texorpdfstring{\href{premv@iucaa.in}{Premvijay Velmani}}{Premvijay Velmani}}
\addresses  {\groupname\\\deptname\\\univname}  % Do not change this here, instead these must be set in the "Thesis.cls" file, please look through it instead
\date       {\today}
\subject    {}
\keywords   {}

\maketitle
%% ----------------------------------------------------------------

\setstretch{1.5}  % It is better to have smaller font and larger line spacing than the other way round

% Define the page headers using the FancyHdr package and set up for one-sided printing
\fancyhead{}  % Clears all page headers and footers
\rhead{\thepage}  % Sets the right side header to show the page number
\lhead{}  % Clears the left side page header

\pagestyle{fancy}  % Finally, use the "fancy" page style to implement the FancyHdr headers

%% ----------------------------------------------------------------
% Declaration Page required for the Thesis, your institution may give you a different text to place here
\Declaration{

\addtocontents{toc}{\vspace{1em}}  % Add a gap in the Contents, for aesthetics

I, Premvijay Velmani, declare that this thesis titled, `Interplay of galaxy formation and the evolution of dark matter haloes in the cosmic web' and the work presented in it are my own. I confirm that:

\begin{itemize} 
\item[\tiny{$\blacksquare$}] This work was done wholly or mainly while in candidature for a research degree at this University.
 
\item[\tiny{$\blacksquare$}] Where any part of this thesis has previously been submitted for a degree or any other qualification at this University or any other institution, this has been clearly stated.
 
\item[\tiny{$\blacksquare$}] Where I have consulted the published work of others, this is always clearly attributed.
 
\item[\tiny{$\blacksquare$}] Where I have quoted from the work of others, the source is always given. With the exception of such quotations, this thesis is entirely my own work.
 
\item[\tiny{$\blacksquare$}] I have acknowledged all main sources of help.
 
\item[\tiny{$\blacksquare$}] Where the thesis is based on work done by myself jointly with others, I have made clear exactly what was done by others and what I have contributed myself.
\\
\end{itemize}
 
 
Signed:\\
\rule[1em]{25em}{0.5pt}  % This prints a line for the signature
 
Date:\\
\rule[1em]{25em}{0.5pt}  % This prints a line to write the date
}
\clearpage  % Declaration ended, now start a new page

%% ----------------------------------------------------------------
% The "Funny Quote Page"
\pagestyle{empty}  % No headers or footers for the following pages

\null\vfill
% Now comes the " Quote", written in italics
% \textit{``Write a quote here.''}

\begin{flushright}
If the quote is taken from someone, their name goes here
\end{flushright}

\vfill\vfill\vfill\vfill\vfill\vfill\null
\clearpage  % Funny Quote page ended, start a new page
%% ----------------------------------------------------------------

% The Abstract Page
\addtotoc{Abstract}  % Add the "Abstract" page entry to the Contents
\abstract{
\addtocontents{toc}{\vspace{1em}}  % Add a gap in the Contents, for aesthetics

Dark matter haloes are gravitationally collapsed structures hosting the formation and evolution of galaxies. While being crucial probes in both cosmology and particle physics, they have often been modeled in gravity-only scenarios, neglecting the effects of baryonic astrophysics. In this thesis, we study the interplay between galaxy formation and its host halo evolution, focusing on changes in the radial distribution of dark matter in response to galaxy formation and evolution. We build a comprehensive understanding using both hydrodynamical simulations of realistic galaxies in cosmological volumes and more tractable self-similarly evolving systems of individual galaxies with their host haloes. This unified picture of the role of various galactic astrophysical processes in mediating the response of dark matter haloes also provides useful timescales for predicting dynamical changes in dark matter mass profiles due to those astrophysical processes.

In the first part of the thesis, we characterise this response statistically in a variety of haloes spanning over four orders of magnitude in mass in the cosmological simulations of galaxy formation from the IllustrisTNG and EAGLE suite at present epoch. We present simple fitting functions in the quasi-adiabatic relaxation framework that accurately capture the dark matter response over the full range of halo mass and halo-centric distance we explore. We show that commonly employed schemes, which consider the relative change in radius $r_f/r_i-1$ of a spherical dark matter shell to be a function of only the relative change in its enclosed mass $M_i/M_f-1$, do not accurately describe the measured response of most haloes in IllustrisTNG and EAGLE. Rather, $r_f/r_i$ additionally explicitly depends upon halo-centric distance $r_f/R_{\rm vir}$ for haloes with virial radius $R_{\rm vir}$, being very similar between IllustrisTNG and EAGLE and across halo mass. We also account for a previously unmodelled effect, likely driven by feedback-related outflows, in which shells having $r_f/r_i\simeq1$ (i.e., no relaxation) have $M_i/M_f$ significantly different from unity.  This difference is usually stronger among the less concentrated recently formed haloes than the older haloes. Similarly, this effect is also prominent in haloes with currently strong star formation activity. These results presented in chapter \ref{chap:z0_main} are immediately applicable to a number of semi-analytical tools for modelling galactic and large-scale structure. We conclude this chapter with a discussion on the possible extensions of these results to build a deeper physical understanding of the connection between dark matter and baryons within haloes.

In the second part, we investigate the role of astrophysical modeling in producing the halo response at different epochs in simulations. We find that the quasi-adaiabtic relaxation with additional dependence on the halo-centric distance is a good description also at earlier epoch. Also the parameters in this model are found to be more Universal across much larger variety of haloes at \(z=1\) than the \(z=0\). Through systematic analysis of a large collection of simulations from the CAMELS project, we find that the baryonic prespriptions for both AGN and stellar feedbacks have a strong influence on the relaxation response of the dark matter halo. In particular, only the parameters controlling the overall feedback energy flux had an effect on the relaxation response, while the wind speed and burstiness have negligible effect on the relaxation at fixed amount of energy flux. However, the exact role of these parameters on the relaxation depends on the redshift. Along with these results in \ref{chap:physvar_z01}, we also study the role of a variety of baryonic astrophysical processes through the EAGLE physics variation simulations. While this depict a similar picture regarding the importance of feedback effects, it also reveals that the gas equation of state to have one of the strongest influence on the relaxation response.

We then present the dynamical evolution of the dark matter's relaxation response to galaxies and their connection to the astrophysical properties as simulated in the IllustrisTNG suite of cosmological hydrodynamical simulations in chapter \ref{chap:dynam-relxn}. Our results show that the radially-dependent linear relaxation relation model proposed in chapter \ref{chap:z0_main} is applicable at least from redshift \(z=5\). We focus on the offset parameter \(q_0\), which characterizes the relaxation of dark matter shells without changing the enclosed mass. We perform multiple time-series analyses to determine the possible causal connections between the relaxation mechanism and astrophysical processes such as star formation and associated feedback processes, as well as feedback due to active galactic nuclei. We show that star formation activity significantly influences the halo relaxation response throughout its evolutionary history, with essentially immediate effects in the inner haloes and delayed effects of 2 to 3 Gyr in the outer regions. Metal content shows a weaker connection to relaxation than star formation rates, but the accumulated wind from feedback processes exhibits a stronger correlation. These findings enhance our understanding of halo relaxation mechanisms. Our estimates of the time-scales relevant for dark matter relaxation can potentially improve the description of halo profiles in existing baryonification schemes and semi-analytical galaxy formation models. Our results also show how the relaxation response of dark haloes can probe the evolutionary history of the galaxies they host.

In the last part of this thesis, we develop a spherical self-similar model for the formation of a galaxy through gas collapsing in an isolated self-gravitating dark matter halo. As is well known, the self-similarity assumption makes the problem eminently tractable by reducing it to a system of ordinary differential equations. We improve upon the existing literature on self-similar collapse in two ways. First, we include the effects of radiative cooling and the formation of a pseudo-disk at the center of collapse, in a parametrised manner. More importantly, we solve for the evolution of gas and dark matter \emph{simultaneously and self-consistently} using a novel iterative approach. As a result, our model produces shell trajectories of both gas \emph{and} dark matter that qualitatively agree with the results of full hydrodynamical simulations of self-gravitating systems. We discuss the impact of various ingredients such as the accretion rate, gas equation of state, disk radius and cooling rate amplitude on the evolution of the gas shells. The self-consistent evolution of gas and dark matter allows us to study the relaxation response of the dark matter trajectories to the presence of collapsing gas. This model produces relaxation relations that are in qualitative agreement with more realistic cosmological hydrodynamical simulations, while allowing us to easily study the impact of the model ingredients mentioned above. As an initial application, we vary one ingredient at a time and find that the accretion rate and gas equation of state have the largest impact on the relaxation relation, while the cooling amplitude plays only a minor role. Our model thus provides a convenient framework to rapidly explore the coupled nonlinear impact of multiple astrophysical processes on the mass and velocity profiles of dark matter in galactic halos, and consequently on observables such as rotation curves and gravitational lensing signals.

% In summary, this thesis presents a comprehensive analysis of the dynamical evolution of the dark matter radial distribution in connection to various baryonic astrophysical processes within haloes. We find that this response is quasi-adiabatic

}

\clearpage  % Abstract ended, start a new page
%% ----------------------------------------------------------------

\setstretch{1.5}  % Reset the line-spacing to 1.3 for body text (if it has changed)

% The Acknowledgements page, for thanking everyone
\acknowledgements{
\addtocontents{toc}{\vspace{1em}}  % Add a gap in the Contents, for aesthetics

Firstly, I would like to acknowledge

}
\clearpage  % End of the Acknowledgements
%% ----------------------------------------------------------------

\pagestyle{fancy}  %The page style headers have been "empty" all this time, now use the "fancy" headers as defined before to bring them back


%% ----------------------------------------------------------------
\lhead{\emph{Contents}}  % Set the left side page header to "Contents"
\tableofcontents  % Write out the Table of Contents

%% ----------------------------------------------------------------
\lhead{\emph{List of Figures}}  % Set the left side page header to "List if Figures"
\listoffigures  % Write out the List of Figures

%% ----------------------------------------------------------------
\lhead{\emph{List of Tables}}  % Set the left side page header to "List of Tables"
\listoftables  % Write out the List of Tables

%% ----------------------------------------------------------------
\setstretch{1.5}  % Set the line spacing to 1.5, this makes the following tables easier to read
\clearpage  % Start a new page
\lhead{\emph{Abbreviations}}  % Set the left side page header to "Abbreviations"
\listofsymbols{ll}  % Include a list of Abbreviations (a table of two columns)
{
% \textbf{Acronym} & \textbf{W}hat (it) \textbf{S}tands \textbf{F}or \\
\textbf{LAH} & \textbf{L}ist \textbf{A}bbreviations \textbf{H}ere \\

}

%% ----------------------------------------------------------------
\clearpage  % Start a new page
\lhead{\emph{Physical Constants}}  % Set the left side page header to "Physical Constants"
\listofconstants{lrcl}  % Include a list of Physical Constants (a four column table)
{
% Constant Name & Symbol & = & Constant Value (with units) \\
Speed of Light & $c$ & $=$ & $2.997\ 924\ 58\times10^{8}\ \mbox{ms}^{-\mbox{s}}$ (exact)\\

}

%% ----------------------------------------------------------------
\clearpage  %Start a new page
\lhead{\emph{Symbols}}  % Set the left side page header to "Symbols"
\listofnomenclature{lll}  % Include a list of Symbols (a three column table)
{
% symbol & name & unit \\
$a$ & distance & m \\
$P$ & power & W (Js$^{-1}$) \\
& & \\ % Gap to separate the Roman symbols from the Greek
$\omega$ & angular frequency & rads$^{-1}$ \\
}
%% ----------------------------------------------------------------
% End of the pre-able, contents and lists of things
% Begin the Dedication page

\setstretch{1.5}  % Return the line spacing back to 1.3

\pagestyle{empty}  % Page style needs to be empty for this page
\dedicatory{For/Dedicated to/To my\ldots}

\addtocontents{toc}{\vspace{2em}}  % Add a gap in the Contents, for aesthetics


%% ----------------------------------------------------------------
\mainmatter	  % Begin normal, numeric (1,2,3...) page numbering
\pagestyle{fancy}  % Return the page headers back to the "fancy" style

% Include the chapters of the thesis, as separate files
% Just uncomment the lines as you write the chapters

\input{Chapters/intro.tex} % Introduction

\input{Chapters/sims_hals.tex}

\input{Chapters/z0_main.tex} % 

\input{Chapters/physvar_z01.tex} % 

\input{Chapters/dynam.tex} % 

\input{Chapters/self_simil.tex} % 

%\input{Chapters/Chapter6} % Results and Discussion

\input{Chapters/concl.tex} % Conclusion

%% ----------------------------------------------------------------
% Now begin the Appendices, including them as separate files

\addtocontents{toc}{\vspace{2em}} % Add a gap in the Contents, for aesthetics

\appendix % Cue to tell LaTeX that the following 'chapters' are Appendices

% \input{Appendices/AppendixA}	% Appendix Title

%\input{Appendices/AppendixB} % Appendix Title

%\input{Appendices/AppendixC} % Appendix Title

\addtocontents{toc}{\vspace{2em}}  % Add a gap in the Contents, for aesthetics
\backmatter

%% ----------------------------------------------------------------
\label{Bibliography}
\lhead{\emph{Bibliography}}  % Change the left side page header to "Bibliography"
\bibliographystyle{unsrtnat}  % Use the "unsrtnat" BibTeX style for formatting the Bibliography
\bibliography{Bibliography,references-AdiabRelxn1,references-SelfSimRelxn}  % The references (bibliography) information are stored in the file named "Bibliography.bib"
% \bibliography{references-AdiabRelxn1}
% \bibliography{references-SelfSimRelxn}
\end{document}  % The End
%% ----------------------------------------------------------------